---
title: "Evaluating Missing Data Handling Techniques Under Different Missingness Mechanisms"
author: 
- Rophence Ojiambo
- Kiran Nagdeo
- Cindy Patippe
geometry: top=1.54cm, bottom=1.94cm, left=1.54cm, right=1.54cm
format:
  pdf:
    include-in-header:
      text: |
        \usepackage{amsmath} 
        \usepackage{fvextra}
        \usepackage{pdflscape}     
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
link-citations: true
colorlinks: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
csl: apa-numeric-superscript-brackets.csl
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE, 
                      dpi = 300)
```


# Introduction


### Diagnosing MCAR: Little’s Test

To evaluate whether the missingness introduced in the simulated dataset is consistent with the Missing Completely At Random (MCAR) assumption, we used Little’s MCAR test implemented in the `naniar` and `misty` R packages. Little’s test is a statistical hypothesis test that examines whether the means of variables with missing data differ significantly between observed and missing groups. The null hypothesis ($H_0$) is that the data are MCAR. The test statistic is calculated as:

$$
\chi^2 = \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}
$$

Where: $O_i$ = Observed frequency and $E_i$ = Expected frequency under the MCAR assumption

In our dataset, the test returned the following values:$\chi^2 = 115.28,\quad \text{df} = 107,\quad p = 0.275$. The number of unique missing data patterns was 8, and 890 out of 2111 observations had at least one missing value. Since the $p$-value is greater than 0.05, we fail to reject the null hypothesis. This suggests that the missing data mechanism is consistent with MCAR. Therefore, we can proceed with analysis methods that assume MCAR without introducing significant bias.

```{r}
# ===============================
# Load Required Libraries
# ===============================
library(mice)        # Multiple imputation using chained equations (MICE)
library(caret)       # Data splitting, model training, and performance metrics
library(pROC)        # ROC curve analysis and AUC calculations
library(tidyverse)   # Data manipulation and visualization
library(naniar)      # Missing data visualization and diagnostics
library(fairml)      # Accessing the obesity.levels dataset used in this study
library(misty)       # Additional missing data analysis tools and diagnostics
library(mvnmle)      # Maximum likelihood estimation with missing values
library(ggplot2)     # For creating publication-quality visualizations
library(gridExtra)   # For arranging multiple ggplot2 plots in a grid layout
library(kableExtra)  # For formatting tables

```


```{r}
# ==============================================================================
# Load and Prepare the Dataset
# ==============================================================================
data(obesity.levels)  # Load dataset from fairml package

# Create binary obesity classification
obesity.levels$obese_binary <- ifelse(obesity.levels$NObeyesdad %in% 
                                      c("Obesity_Type_I", "Obesity_Type_II", "Obesity_Type_III"), 
                                      1, 0)
obesity.levels$obese_binary <- factor(obesity.levels$obese_binary, 
                                     levels = c(0, 1), 
                                     labels = c("Not Obese", "Obese"))

# Define variable types
cont_vars <- c("Age", "FCVC", "NCP", "CH2O", "FAF", "TUE")  
cat_vars <- c("Gender", "family_history_with_overweight", "FAVC", "CAEC", 
              "SMOKE", "SCC", "CALC", "MTRANS")  
target_var <- "obese_binary"

# Final dataset without Height/Weight
obesity_full <- obesity.levels %>% 
  select(all_of(c(cont_vars, cat_vars, target_var)))

# ==============================================================================
# Improved Missing Data Simulation Functions
# ==============================================================================

# MCAR: Missing completely at random
simulate_mcar <- function(data, vars, rates) {
  set.seed(123)
  mcar_data <- data
  for (i in seq_along(vars)) {
    var <- vars[i]
    rate <- rates[i]
    missing_idx <- sample(nrow(data), size = floor(rate * nrow(data)))
    mcar_data[missing_idx, var] <- NA
  }
  return(mcar_data)
}

# MAR: Missing at random (depends on other observed variables)
simulate_mar <- function(data) {
  set.seed(123)
  mar_data <- data
  
  # FAF missingness: Higher in older males
  age_scaled <- scale(data$Age)
  male <- as.numeric(data$Gender == "Male")
  logit_faf <- -1.5 + 0.8*age_scaled + 0.7*male
  p_faf <- plogis(logit_faf)
  missing_faf <- rbinom(nrow(data), 1, p_faf) == 1
  mar_data$FAF[missing_faf] <- NA
  
  # SCC missingness: Higher in those who consume high-calorie food
  favc_yes <- as.numeric(data$FAVC == "yes")
  logit_scc <- -2.0 + 1.2*favc_yes
  p_scc <- plogis(logit_scc)
  missing_scc <- rbinom(nrow(data), 1, p_scc) == 1
  mar_data$SCC[missing_scc] <- NA
  
  # NCP missingness: Higher in younger females
  female <- as.numeric(data$Gender == "Female")
  age_scaled <- scale(data$Age)
  logit_ncp <- -1.0 - 0.9*age_scaled + 0.8*female
  p_ncp <- plogis(logit_ncp)
  missing_ncp <- rbinom(nrow(data), 1, p_ncp) == 1
  mar_data$NCP[missing_ncp] <- NA
  
  return(mar_data)
}

# MNAR: Missing not at random (depends on unobserved values)
simulate_mnar <- function(data) {
  set.seed(123)
  mnar_data <- data
  
  # FAF missingness: Higher in inactive people
  faf_low <- as.numeric(data$FAF < 1)
  logit_faf <- -1.5 + 1.8*faf_low
  p_faf <- plogis(logit_faf)
  missing_faf <- rbinom(nrow(data), 1, p_faf) == 1
  mnar_data$FAF[missing_faf] <- NA
  
  # CALC missingness: Higher in frequent drinkers
  calc_freq <- as.numeric(data$CALC == "Frequently")
  logit_calc <- -2.0 + 1.5*calc_freq
  p_calc <- plogis(logit_calc)
  missing_calc <- rbinom(nrow(data), 1, p_calc) == 1
  mnar_data$CALC[missing_calc] <- NA
  
  # SCC missingness: Higher in those who don't monitor calories
  scc_no <- as.numeric(data$SCC == "no")
  logit_scc <- -1.8 + 1.4*scc_no
  p_scc <- plogis(logit_scc)
  missing_scc <- rbinom(nrow(data), 1, p_scc) == 1
  mnar_data$SCC[missing_scc] <- NA
  
  return(mnar_data)
}

# Apply missingness mechanisms
mcar_data <- simulate_mcar(obesity_full, 
                          vars = c("Gender", "FAF", "CH2O"), 
                          rates = c(0.15, 0.25, 0.20))

mar_data <- simulate_mar(obesity_full)
mnar_data <- simulate_mnar(obesity_full)
```


```{r}
# ==============================================================================
# Apply Imputation Methods
# ==============================================================================

# Function to apply mean/mode imputation
mean_mode_impute <- function(data) {
  imp_data <- data
  # Impute continuous variables with mean
  for (var in cont_vars) {
    if (any(is.na(imp_data[[var]]))) {
      imp_data[is.na(imp_data[[var]]), var] <- mean(imp_data[[var]], na.rm = TRUE)
    }
  }
  # Impute categorical variables with mode
  for (var in cat_vars) {
    if (any(is.na(imp_data[[var]]))) {
      mode_val <- names(sort(table(imp_data[[var]]), decreasing = TRUE))[1]
      imp_data[is.na(imp_data[[var]]), var] <- mode_val
    }
  }
  return(imp_data)
}

# Apply imputation to all scenarios
impute_all <- function(scenario_data) {
  imputed <- list()
  
  # Complete case
  imputed$complete_case <- na.omit(scenario_data)
  
  # Mean/mode imputation
  imputed$mean_mode <- mean_mode_impute(scenario_data)
  
  # MICE imputation (5 imputations)
  mice_imp <- mice(scenario_data, m = 5, maxit = 10, printFlag = FALSE)
  imputed$mice <- lapply(1:5, function(i) complete(mice_imp, i))
  
  return(imputed)
}

# Apply to all scenarios
mcar_imputed <- impute_all(mcar_data)
mar_imputed <- impute_all(mar_data)
mnar_imputed <- impute_all(mnar_data)
```


```{r}
# ==============================================================================
# Data Splitting Function
# ==============================================================================
split_dataset <- function(data, target = "obese_binary", train_frac = 0.7) {
  set.seed(42)
  # Ensure both classes are present
  idx <- createDataPartition(data[[target]], p = train_frac, list = FALSE)
  train <- data[idx, ]
  test  <- data[-idx, ]
  
  # Check if both classes exist
  if (length(unique(train[[target]])) < 2 | length(unique(test[[target]])) < 2) {
    # Re-sample if any set has only one class
    idx <- createDataPartition(data[[target]], p = train_frac, list = FALSE)
    train <- data[idx, ]
    test  <- data[-idx, ]
  }
  
  return(list(train = train, test = test))
}

# Split all datasets
splits <- list(
  full = split_dataset(obesity_full),
  
  mcar = list(
    original = split_dataset(mcar_data),
    complete_case = split_dataset(mcar_imputed$complete_case),
    mean_mode = split_dataset(mcar_imputed$mean_mode),
    mice = lapply(mcar_imputed$mice, split_dataset)
  ),
  
  mar = list(
    original = split_dataset(mar_data),
    complete_case = split_dataset(mar_imputed$complete_case),
    mean_mode = split_dataset(mar_imputed$mean_mode),
    mice = lapply(mar_imputed$mice, split_dataset)
  ),
  
  mnar = list(
    original = split_dataset(mnar_data),
    complete_case = split_dataset(mnar_imputed$complete_case),
    mean_mode = split_dataset(mnar_imputed$mean_mode),
    mice = lapply(mnar_imputed$mice, split_dataset)
  )
)
```


```{r}
# ==============================================================================
# Performance Evaluation Functions 
# ==============================================================================

# Corrected main performance function
run_logistic_performance <- function(train_data, test_data) {
  # Convert categorical to factors
  train_data <- train_data %>% mutate(across(all_of(cat_vars), as.factor))
  test_data <- test_data %>% mutate(across(all_of(cat_vars), as.factor))
  
  # Fit model
  model <- glm(obese_binary ~ ., data = train_data, family = binomial)
  
  # Calculate metrics using corrected function
  train_metrics <- calculate_metrics(model, train_data)
  test_metrics <- calculate_metrics(model, test_data)
  
  return(list(train = train_metrics, test = test_metrics))
}

# Metrics calculation function
calculate_metrics <- function(model, data) {
  prob_pred <- predict(model, newdata = data, type = "response")
  class_pred <- ifelse(prob_pred > 0.5, "Obese", "Not Obese")
  class_pred <- factor(class_pred, levels = c("Not Obese", "Obese"))
  
  # Handle cases where predictions don't cover all classes
  if (length(unique(class_pred)) < 2) {
    return(list(
      Accuracy = NA, Sensitivity = NA, 
      Specificity = NA, AUC = NA, F1 = NA
    ))
  }
  
  cm <- confusionMatrix(class_pred, data$obese_binary, positive = "Obese")
  auc_val <- auc(roc(data$obese_binary, prob_pred))
  
  return(list(
    Accuracy = as.numeric(cm$overall["Accuracy"]),
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    AUC = as.numeric(auc_val),
    F1 = as.numeric(cm$byClass["F1"])
  ))
}

# MICE-specific processing with Rubin's rules
process_mice_performance <- function(mice_splits) {
  models <- list()
  test_preds <- list()
  
  # Train models and collect test predictions
  for (i in 1:length(mice_splits)) {
    split <- mice_splits[[i]]
    
    # Convert categoricals to factors
    split$train <- split$train %>% mutate(across(all_of(cat_vars), as.factor))
    split$test <- split$test %>% mutate(across(all_of(cat_vars), as.factor))
    
    model <- glm(obese_binary ~ ., data = split$train, family = binomial)
    models[[i]] <- model
    
    # Get test predictions
    test_preds[[i]] <- predict(model, newdata = split$test, type = "response")
  }
  
  # Pool predictions using Rubin's rules
  pooled_preds <- rowMeans(do.call(cbind, test_preds))
  
  # Calculate metrics using pooled predictions
  calculate_pooled_metrics <- function(preds, data) {
    class_pred <- ifelse(preds > 0.5, "Obese", "Not Obese")
    class_pred <- factor(class_pred, levels = c("Not Obese", "Obese"))
    
    if (nlevels(class_pred) < 2) {
      return(list(
        Accuracy = NA, Sensitivity = NA, 
        Specificity = NA, AUC = NA, F1 = NA
      ))
    }
    
    cm <- confusionMatrix(class_pred, data$obese_binary, positive = "Obese")
    auc_val <- auc(roc(data$obese_binary, preds))
    
    return(list(
      Accuracy = as.numeric(cm$overall["Accuracy"]),
      Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
      Specificity = as.numeric(cm$byClass["Specificity"]),
      AUC = as.numeric(auc_val),
      F1 = as.numeric(cm$byClass["F1"])
    ))
  }
  
  # Use first test set as reference
  test_ref <- mice_splits[[1]]$test
  test_metrics <- calculate_pooled_metrics(pooled_preds, test_ref)
  
  # For training, average metrics across imputations
  train_metrics_list <- map(models, ~ calculate_metrics(.x, .x$data))
  train_metrics <- list(
    Accuracy = mean(map_dbl(train_metrics_list, "Accuracy"), na.rm = TRUE),
    Sensitivity = mean(map_dbl(train_metrics_list, "Sensitivity"), na.rm = TRUE),
    Specificity = mean(map_dbl(train_metrics_list, "Specificity"), na.rm = TRUE),
    AUC = mean(map_dbl(train_metrics_list, "AUC"), na.rm = TRUE),
    F1 = mean(map_dbl(train_metrics_list, "F1"), na.rm = TRUE)
  )
  
  return(list(train = train_metrics, test = test_metrics))
}
```


```{r}
# ==============================================================================
# Evaluate All Scenarios 
# ==============================================================================
results <- list()

# Full dataset baseline
results$full <- run_logistic_performance(splits$full$train, splits$full$test)

# Evaluate scenarios
scenarios <- c("mcar", "mar", "mnar")
methods <- c("complete_case", "mean_mode", "mice")

for (scenario in scenarios) {
  scenario_results <- list()
  
  # Original missing data
  scenario_results$original <- run_logistic_performance(
    splits[[scenario]]$original$train,
    splits[[scenario]]$original$test
  )
  
  # Imputation methods
  for (method in methods) {
    if (method == "mice") {
      # MICE requires special handling
      scenario_results[[method]] <- process_mice_performance(splits[[scenario]][[method]])
    } else {
      # Other methods
      split_data <- splits[[scenario]][[method]]
      scenario_results[[method]] <- run_logistic_performance(
        train_data = split_data$train,
        test_data = split_data$test
      )
    }
  }
  
  results[[scenario]] <- scenario_results
}

```



```{r}
# ==============================================================================
# Create Performance Tables
# ==============================================================================
create_performance_table <- function(results, dataset = "test") {
  # Initialize data frame
  perf_df <- data.frame(
    Scenario = character(),
    Method = character(),
    Accuracy = numeric(),
    Sensitivity = numeric(),
    Specificity = numeric(),
    AUC = numeric(),
    F1 = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Add full dataset results
  perf_df <- rbind(perf_df, data.frame(
    Scenario = "Full",
    Method = "Full Dataset",
    Accuracy = results$full[[dataset]]$Accuracy,
    Sensitivity = results$full[[dataset]]$Sensitivity,
    Specificity = results$full[[dataset]]$Specificity,
    AUC = results$full[[dataset]]$AUC,
    F1 = results$full[[dataset]]$F1
  ))
  
  # Add scenario results
  for (scenario in c("mcar", "mar", "mnar")) {
    # Original missing data
    perf_df <- rbind(perf_df, data.frame(
      Scenario = toupper(scenario),
      Method = "original",
      Accuracy = results[[scenario]]$original[[dataset]]$Accuracy,
      Sensitivity = results[[scenario]]$original[[dataset]]$Sensitivity,
      Specificity = results[[scenario]]$original[[dataset]]$Specificity,
      AUC = results[[scenario]]$original[[dataset]]$AUC,
      F1 = results[[scenario]]$original[[dataset]]$F1
    ))
    
    # Imputation methods
    for (method in methods) {
      perf_df <- rbind(perf_df, data.frame(
        Scenario = toupper(scenario),
        Method = method,
        Accuracy = results[[scenario]][[method]][[dataset]]$Accuracy,
        Sensitivity = results[[scenario]][[method]][[dataset]]$Sensitivity,
        Specificity = results[[scenario]][[method]][[dataset]]$Specificity,
        AUC = results[[scenario]][[method]][[dataset]]$AUC,
        F1 = results[[scenario]][[method]][[dataset]]$F1
      ))
    }
  }
  
  # Round values
  perf_df[, 3:7] <- round(perf_df[, 3:7], 4)
  
  return(perf_df)
}

# Create tables
training_table <- create_performance_table(results, "train")
test_table <- create_performance_table(results, "test")

# Format tables
training_table <- training_table %>%
  rename(`F1 Score` = F1) %>%
  select(Scenario, Method, Accuracy, Sensitivity, Specificity, AUC, `F1 Score`)

test_table <- test_table %>%
  rename(`F1 Score` = F1) %>%
  select(Scenario, Method, Accuracy, Sensitivity, Specificity, AUC, `F1 Score`)
```



```{r}
training_table <- training_table %>%
  mutate(Method = case_when(
      Method == "original" ~ "Simulated (with missing)",
      Method == "complete_case" ~ "Complete Case",
      Method == "mean_mode" ~ "Mean/Mode Imputation",
      Method == "mice" ~ "Multiple Imputation (MICE)",
      Method == "Full Dataset" ~ "Original Dataset (No Missing)",
      TRUE ~ Method))

# Displaying results
training_table %>%
  kbl(booktabs = TRUE, digits = 5, caption = "Model Performance on Training Set Across Missing Data Handling Methods") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  group_rows("Full Dataset", 1, 1) %>%
  group_rows("MCAR", 2, 5) %>%
  group_rows("MAR", 6, 9) %>%
  group_rows("MNAR", 10, 13)
```


```{r}
test_table <- test_table %>%
  mutate(Method = case_when(
      Method == "original" ~ "Simulated (with missing)",
      Method == "complete_case" ~ "Complete Case",
      Method == "mean_mode" ~ "Mean/Mode Imputation",
      Method == "mice" ~ "Multiple Imputation (MICE)",
      Method == "Full Dataset" ~ "Original Dataset (No Missing)",
      TRUE ~ Method))

# Displaying results
test_table %>%
  kbl(booktabs = TRUE, digits = 5, caption = "Model Performance on Test Set Across Missing Data Handling Methods") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  group_rows("Full Dataset", 1, 1) %>%
  group_rows("MCAR", 2, 5) %>%
  group_rows("MAR", 6, 9) %>%
  group_rows("MNAR", 10, 13)
```



```{r}
# ==============================================================================
# Diagnosing MCAR: Little’s Test
# ==============================================================================
# Perform Little's MCAR test using the `naniar` package
naniar::mcar_test(mcar_data)

# Perform the same test using the `misty` package for additional summary
misty::na.test(mcar_data)

# ==============================================================================
# Visualize MCAR Missingness Patterns
# ==============================================================================

# Heatmap of missing values
vis_miss(mcar_data) + ggtitle("MCAR Missingness Pattern")

# UpSet plot to show combinations of missingness
gg_miss_upset(mcar_data) 

# Visualize missingness across Obesity Levels
gg_miss_fct(x = mcar_data, fct = obese_binary) + 
  labs(title = "MCAR Missingness in Risk Factors and Obesity Level",
       x = "Obesity Levels", y = "Risk Factors")
```



```{r}
# ==============================================================================
# Visualize MAR Missingness Patterns
# ==============================================================================

# Heatmap of missing values
vis_miss(mar_data) + ggtitle("MAR Missingness Pattern")

# UpSet plot to show combinations of missingness
gg_miss_upset(mar_data) 

# Visualize missingness across Obesity Levels
gg_miss_fct(x = mar_data, fct = obese_binary) + 
  labs(title = "MAR Missingness in Risk Factors and Obesity Level",
       x = "Obesity Levels", y = "Risk Factors")
```




```{r}
# ==============================================================================
# Visualize MNAR Missingness Patterns
# ==============================================================================

# Heatmap of missing values
vis_miss(mnar_data) + ggtitle("MNAR Missingness Pattern")

# UpSet plot to show combinations of missingness
gg_miss_upset(mnar_data) 

# Visualize missingness across Obesity Levels
gg_miss_fct(x = mnar_data, fct = obese_binary) + 
  labs(title = "MNAR Missingness in Risk Factors and Obesity Level",
       x = "Obesity Levels", y = "Risk Factors")
```






\newpage

# Data Availabiliy

All codes and datasets for this project can be found at our [GitHub Repository](https://github.com/rophenceojiambo/Simulations-Final-Project).

# References

