---
title: "Evaluating Missing Data Handling Techniques Under Different Missingness Mechanisms"
author: 
- Rophence Ojiambo
- Kiran Nagdeo
- Cindy Patippe
geometry: top=1.54cm, bottom=1.94cm, left=1.54cm, right=1.54cm
format:
  pdf:
    include-in-header:
      text: |
        \usepackage{amsmath} 
        \usepackage{fvextra}
        \usepackage{pdflscape}     
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
link-citations: true
colorlinks: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
csl: apa-numeric-superscript-brackets.csl
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE, 
                      dpi = 300)
```


# Introduction

```{r}
# ===============================
# Load Required Libraries
# ===============================
library(mice)        # For multiple imputation by chained equations (MICE)
library(doParallel)  # For parallel processing (if needed later)
library(foreach)     # For parallel looping
library(purrr)       # For functional programming (e.g., map)
library(fairml)      # For accessing the `obesity.levels` dataset
library(naniar)
library(misty)
library(mvnmle)
library(ggplot2)
library(tidyverse)
```


```{r}
# ==============================================================================
# Load and Prepare the Dataset
# ==============================================================================

data(obesity.levels)  # Load dataset from fairml package

# Continuous variables
cont_vars <- c("Age", "Height", "Weight", "FCVC", "NCP", "CH2O", "FAF", "TUE")  

# Categorical variables
cat_vars <- c("Gender", "family_history_with_overweight", "FAVC", "CAEC", 
              "SMOKE", "SCC", "CALC", "MTRANS")  

target_var <- "NObeyesdad"  # Target variable (obesity classification)

# Subset to relevant columns
obesity_full <- obesity.levels[, c(cont_vars, cat_vars, target_var)]

# ==============================================================================
# Introduce variable-specific MCAR: 4%, 24%, and 7% for selected variables
# ==============================================================================
set.seed(123)  # For reproducibility

mcar_vars <- c("Gender", "FAF", "CH2O")            # Variables to introduce MCAR
mcar_rates <- c(0.07, 0.24, 0.18)                  # Corresponding MCAR rates
mcar_data <- obesity_full                          # Copy of original dataset

# Loop through variables and introduce corresponding amount of MCAR
for (i in seq_along(mcar_vars)) {
  var <- mcar_vars[i]
  rate <- mcar_rates[i]
  
  missing_idx <- sample(nrow(obesity_full), size = floor(rate * nrow(obesity_full)))
  mcar_data[missing_idx, var] <- NA
}


# ==============================================================================
# Apply Imputation Methods
# ==============================================================================

imputed <- list()

# --- (a) Complete Case Analysis ---
imputed$complete_case <- na.omit(mcar_data)

# --- (b) Mean/Mode Imputation ---
mean_imp <- mcar_data

# Impute continuous variables with mean
for (var in cont_vars) {
  mean_imp[is.na(mean_imp[[var]]), var] <- mean(mean_imp[[var]], na.rm = TRUE)
}

# Impute categorical variables with mode
for (var in cat_vars) {
  mode_val <- names(which.max(table(mean_imp[[var]])))
  mean_imp[is.na(mean_imp[[var]]), var] <- mode_val
}

imputed$mean_mode <- mean_imp

# --- (c) Multiple Imputation using MICE ---
mice_imp <- mice(mcar_data, m = 5, maxit = 10, printFlag = FALSE)
imputed$mice <- lapply(1:5, function(i) complete(mice_imp, i))

# ==============================================================================
# Save Results
# ==============================================================================
final_results <- list(mcar = mcar_data, imputed = imputed)

saveRDS(final_results, "mcar_imputation_5mice.rds", compress = "xz")
```


```{r}
# ==============================================================================
# Load Saved Imputation Results
# ==============================================================================
final_results <- readRDS("mcar_imputation_5mice.rds")

# Access individual datasets
mcar_data <- final_results$mcar
mcar_complete_case <- final_results$imputed$complete_case
mcar_mean_mode <- final_results$imputed$mean_mode
mcar_mice <- final_results$imputed$mice

# Extract individual MICE datasets
mcar_mice_1 <- mcar_mice[[1]]
mcar_mice_2 <- mcar_mice[[2]]
mcar_mice_3 <- mcar_mice[[3]]
mcar_mice_4 <- mcar_mice[[4]]
mcar_mice_5 <- mcar_mice[[5]]
```


### Diagnosing MCAR: Little’s Test

To evaluate whether the missingness introduced in the simulated dataset is consistent with the Missing Completely At Random (MCAR) assumption, we used Little’s MCAR test implemented in the `naniar` and `misty` R packages. Little’s test is a statistical hypothesis test that examines whether the means of variables with missing data differ significantly between observed and missing groups. The null hypothesis ($H_0$) is that the data are MCAR. The test statistic is calculated as:

$$
\chi^2 = \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}
$$

Where: $O_i$ = Observed frequency and $E_i$ = Expected frequency under the MCAR assumption

In our dataset, the test returned the following values:$\chi^2 = 115.28,\quad \text{df} = 107,\quad p = 0.275$. The number of unique missing data patterns was 8, and 890 out of 2111 observations had at least one missing value. Since the $p$-value is greater than 0.05, we fail to reject the null hypothesis. This suggests that the missing data mechanism is consistent with MCAR. Therefore, we can proceed with analysis methods that assume MCAR without introducing significant bias.


```{r, results='hide'}
# ==============================================================================
# Diagnosing MCAR: Little’s Test
# ==============================================================================
# Perform Little's MCAR test using the `naniar` package
naniar::mcar_test(mcar_data)

# Perform the same test using the `misty` package for additional summary
misty::na.test(mcar_data)
```

```{r}
# ==============================================================================
# Visualize MCAR Missingness Patterns
# ==============================================================================

# Heatmap of missing values
vis_miss(mcar_data)

# UpSet plot to show combinations of missingness
gg_miss_upset(mcar_data) 

# Visualize missingness across Obesity Levels
gg_miss_fct(x = mcar_data, fct = NObeyesdad) + 
  labs(title = "MCAR Missingness in Risk Factors and Obesity Level",
       x = "Obesity Levels", y = "Risk Factors")
```




```{r}
# ==============================================================================
# Simulate MAR Missingness in Obesity Dataset
# ==============================================================================

set.seed(123)
obesity_mar <- obesity_full

# ==============================================================================
# Missingness in 'FAF' (Physical Activity Frequency)
# Based on: Age and CH2O (both fully observed)
# ==============================================================================

age_scaled <- scale(obesity_full$Age)
ch2o_scaled <- scale(obesity_full$CH2O)

lin_pred_faf <- 0.9 * age_scaled - 0.6 * ch2o_scaled  # Older age ↑ missing; more water ↓ missing
prob_faf_miss <- plogis(lin_pred_faf) * 0.17          # Cap missingness around 17%
missing_idx_faf <- which(runif(nrow(obesity_full)) < prob_faf_miss)
obesity_mar[missing_idx_faf, "FAF"] <- NA

# ==============================================================================
# Missingness in 'SMOKE'
# Based on: family_history_with_overweight and Gender
# ==============================================================================

base_prob <- 0.05
fam_hist_effect <- ifelse(obesity_full$family_history_with_overweight == "yes", 0.04, 0)
gender_effect_smoke <- ifelse(obesity_full$Gender == "Male", 0.08, 0)
prob_smoke_miss <- pmin(base_prob + fam_hist_effect + gender_effect_smoke, 0.21)  # Max 21%

missing_idx_smoke <- which(runif(nrow(obesity_full)) < prob_smoke_miss)
obesity_mar[missing_idx_smoke, "SMOKE"] <- NA

# ==============================================================================
# Missingness in 'SCC' (Calorie Consumption Monitoring)
# Based on: FAVC (frequent high-calorie food) and Gender
# ==============================================================================

base_prob <- 0.03
favc_effect <- ifelse(obesity_full$FAVC == "yes", 0.04, 0)
gender_effect_scc <- ifelse(obesity_full$Gender == "Male", 0.05, 0)
prob_scc_miss <- pmin(base_prob + favc_effect + gender_effect_scc, 0.16) # Max 16%

missing_idx_scc <- which(runif(nrow(obesity_full)) < prob_scc_miss)
obesity_mar[missing_idx_scc, "SCC"] <- NA

# ==============================================================================
# Missingness in 'NCP' (Number of Main Meals)
# Based on: Age and Gender
# ==============================================================================

gender_female <- ifelse(obesity_full$Gender == "Female", 0.2, 0)
lin_pred_ncp <- -0.8 * age_scaled + gender_female  # Younger and female ↑ missingness
prob_missing_ncp <- plogis(lin_pred_ncp) * 0.2     # Cap at 20%

missing_idx_ncp <- which(runif(nrow(obesity_full)) < prob_missing_ncp)
obesity_mar[missing_idx_ncp, "NCP"] <- NA

# ==============================================================================
# Apply Imputation Methods
# ==============================================================================

mar_imputed <- list()

# --- (a) Complete Case Analysis ---
mar_imputed$mar_complete_case <- na.omit(obesity_mar)

# --- (b) Mean/Mode Imputation ---
mar_mean_imp <- obesity_mar

# Impute continuous variables with mean
for (var in cont_vars) {
  mar_mean_imp[is.na(mar_mean_imp[[var]]), var] <- mean(mar_mean_imp[[var]], na.rm = TRUE)
}

# Impute categorical variables with mode
for (var in cat_vars) {
  mar_mode_val <- names(which.max(table(mar_mean_imp[[var]])))
  mar_mean_imp[is.na(mar_mean_imp[[var]]), var] <- mar_mode_val
}

mar_imputed$mar_mean_mode <- mar_mean_imp

# --- (c) Multiple Imputation using MICE ---
mar_mice_imp <- mice(obesity_mar, m = 5, maxit = 10, printFlag = FALSE)
mar_imputed$mar_mice <- lapply(1:5, function(i) complete(mar_mice_imp, i))

# ==============================================================================
# Save Results
# ==============================================================================
mar_results <- list(mar = obesity_mar, imputed = mar_imputed)

saveRDS(mar_results, "mar_imputation_5mice.rds", compress = "xz")

# ==============================================================================
# Load Saved Imputation Results
# ==============================================================================
mar_results <- readRDS("mar_imputation_5mice.rds")

# Access individual datasets
mar_data <- mar_results$mar
mar_complete_case <- mar_results$imputed$complete_case
mar_mean_mode <- mar_results$imputed$mar_mean_mode
mar_mice <- mar_results$imputed$mar_mice

# Extract individual MICE datasets
mar_mice_1 <- mar_mice[[1]]
mar_mice_2 <- mar_mice[[2]]
mar_mice_3 <- mar_mice[[3]]
mar_mice_4 <- mar_mice[[4]]
mar_mice_5 <- mar_mice[[5]]
```


```{r}
# ==============================================================================
# Visualize MAR Missingness Patterns
# ==============================================================================

# Heatmap of missing values
vis_miss(mar_data)

# UpSet plot to show combinations of missingness
gg_miss_upset(mar_data) 

# Visualize missingness across Obesity Levels
gg_miss_fct(x = mar_data, fct = NObeyesdad) + 
  labs(title = "MAR Missingness in Risk Factors and Obesity Level",
       x = "Obesity Levels", y = "Risk Factors")
```




```{r}
# ==============================================================================
# Simulate MNAR Missingness in Obesity Dataset
# ==============================================================================

set.seed(123)  # For reproducibility

# Make a copy of the complete dataset
obesity_mnar <- obesity_full

# --- (1) SCC (Calories Monitoring) ---
# Simulate MNAR: Missingness among those who answered "no" to SCC
idx_scc_no <- which(obesity_mnar$SCC == "no")
miss_scc <- sample(idx_scc_no, size = round(0.3 * length(idx_scc_no)))  # 30% of "no"
obesity_mnar$SCC[miss_scc] <- NA

# --- (2) FAF (Physical Activity Level) ---
# Simulate MNAR: People with very low physical activity (FAF < 1)
idx_faf_low <- which(obesity_mnar$FAF < 1)
miss_faf <- sample(idx_faf_low, size = round(0.4 * length(idx_faf_low)))  # 40% missing
obesity_mnar$FAF[miss_faf] <- NA

# --- (3) CALC (Alcohol Consumption) ---
# Simulate MNAR: Frequent drinkers ("Frequently") more likely to skip
idx_calc_freq <- which(obesity_mnar$CALC == "Frequently")
miss_calc <- sample(idx_calc_freq, size = round(0.5 * length(idx_calc_freq)))  # 50% missing
obesity_mnar$CALC[miss_calc] <- NA

# ==============================================================================
# Apply Imputation Methods for MNAR Dataset
# ==============================================================================

mnar_imputed <- list()

# --- (a) Complete Case Analysis ---
mnar_imputed$mnar_complete_case <- na.omit(obesity_mnar)

# --- (b) Mean/Mode Imputation ---
mnar_mean_imp <- obesity_mnar

# Impute continuous variables with mean
for (var in cont_vars) {
  mnar_mean_imp[is.na(mnar_mean_imp[[var]]), var] <- mean(mnar_mean_imp[[var]], na.rm = TRUE)
}

# Impute categorical variables with mode
for (var in cat_vars) {
  mnar_mode_val <- names(which.max(table(mnar_mean_imp[[var]])))
  mnar_mean_imp[is.na(mnar_mean_imp[[var]]), var] <- mnar_mode_val
}

mnar_imputed$mnar_mean_mode <- mnar_mean_imp

# --- (c) Multiple Imputation using MICE ---
mnar_mice_imp <- mice(obesity_mnar, m = 5, maxit = 10, printFlag = FALSE)
mnar_imputed$mnar_mice <- lapply(1:5, function(i) complete(mnar_mice_imp, i))

# ==============================================================================
# Save Results
# ==============================================================================
mnar_results <- list(mnar = obesity_mnar, imputed = mnar_imputed)

saveRDS(mnar_results, "mnar_imputation_5mice.rds", compress = "xz")


# ==============================================================================
# Load Saved Imputation Results
# ==============================================================================
mnar_results <- readRDS("mnar_imputation_5mice.rds")

# Access individual datasets
mnar_data <- mnar_results$mnar
mnar_complete_case <- mnar_results$imputed$mnar_complete_case
mnar_mean_mode <- mnar_results$imputed$mnar_mean_mode
mnar_mice <- mnar_results$imputed$mnar_mice

# Extract individual MICE datasets
mnar_mice_1 <- mnar_mice[[1]]
mnar_mice_2 <- mnar_mice[[2]]
mnar_mice_3 <- mnar_mice[[3]]
mnar_mice_4 <- mnar_mice[[4]]
mnar_mice_5 <- mnar_mice[[5]]

```


```{r}
# ==============================================================================
# Visualize MNAR Missingness Patterns
# ==============================================================================

# Heatmap of missing values
vis_miss(mnar_data)

# UpSet plot to show combinations of missingness
gg_miss_upset(mnar_data) 

# Visualize missingness across Obesity Levels
gg_miss_fct(x = mnar_data, fct = NObeyesdad) + 
  labs(title = "MNAR Missingness in Risk Factors and Obesity Level",
       x = "Obesity Levels", y = "Risk Factors")
```


\newpage

# Data Availabiliy

All codes and datasets for this project can be found at our [GitHub Repository](https://github.com/rophenceojiambo/Simulations-Final-Project).

# References

